---
title: "Drug Shortages Analysis"
author: "Bryan Paget"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
includes:
  after_body: footer.html
---

<style>
  .blue {
    border-radius: 4px;
    padding: 16px;
    font-size: large;
    font-weight: 800;
  }
</style>

<div class = "blue">

## Question

a) By using the data available on [Drug Shortages Canada](https://www.drugshortagescanada.ca/) and [Drug Product Database (DPD)](https://www.canada.ca/en/health-canada/services/drugs-health-products/drug-products/drug-product-database/what-data-extract-drug-product-database.html), identify the percentage of actual drug shortage reports to Health Canada that did not have a prior report of an anticipated or actual shortage within a 12-month period for the same drug. Restrict your collection of data to the period June 2019 to June 2021.

b) For those drugs that had the type of report that was of concern in a), what was the average age on the market?

</div>

```{r settings-and-import-libraries, message=FALSE, warning=FALSE, include=TRUE, paged.print=FALSE, results='hide'}
library(tidyverse)
library(docstring)
library(magrittr)
library(janitor)
library(lubridate)
library(rdrugshortages)

#' Disable scientific notation

options(scipen = 999)

#' This is the download location of the data we are interested in from the Drug Product Database:

dpd_file_location <- "https://www.canada.ca/content/dam/hc-sc/documents/services/drug-product-database/allfiles.zip"

#' Start and end dates for Drug Shortages Canada Web API:

my_start_date <- ymd(20190601, tz = "America/Toronto")
my_end_date <- ymd(20210601, tz = "America/Toronto")

#' Create date interval object, used for filtering records after we've pulled them using the API:

my_date_range <- interval(start = my_start_date,
                          end = my_end_date,
                          tzone = "America/Toronto")

#' The tables from the Drug Product Database don't come with column names, they are provided here:

coltypes_status <- cols(
  col_double(),
  col_character(),
  col_character(),
  col_character(),
  col_character(),
  col_character(),
  col_character()
)

colnames_status <- c(
  "DRUG_CODE",
  "CURRENT_STATUS_FLAG",
  "STATUS",
  "HISTORY_DATE",
  "STATUS_F",
  "LOT_NUMBER",
  "EXPIRATION_DATE"
)

coltypes_drug <- cols(
  col_double(),
  col_character(),
  col_character(),
  col_character(),
  col_character(),
  col_character(),
  col_character(),
  col_character(),
  col_double(),
  col_character(),
  col_character(),
  col_character(),
  col_character(),
  col_character()
)

colnames_drug <- c(
  "DRUG_CODE",
  "PRODUCT_CATEGORIZATION",
  "CLASS",
  "DRUG_IDENTIFICATION_NUMBER",
  "BRAND_NAME",
  "DESCRIPTOR",
  "PEDIATRIC_FLAG",
  "ACCESSION_NUMBER",
  "NUMBER_OF_AIS",
  "LAST_UPDATE_DATE",
  "AI_GROUP_NO",
  "CLASS_F",
  "BRAND_NAME_F",
  "DESCRIPTOR_F"
)

```

<br>

## Retrieving Data

I'm going to start by pulling data from [Drug Shortages Canada](https://www.drugshortagescanada.ca/) using it's API. For more information about the API, check out [the API's documentation](https://www.drugshortagescanada.ca/blog/52).

There is an `R` package called [`rdrugshortages`](https://github.com/pipitone/rdrugshortages), I'm going to use that. Later I'll filter the data to exclude records added outside the following date range: `r my_start_date` to `r my_end_date`.

So let's go ahead and get some drug shortages data:

```{r import-data, message=TRUE, warning=TRUE, cache=TRUE}
drug_shortages_data <- dsc_search()
```

I'm going to quickly format some columns as dates.

```{r fixup-types}
drug_shortages_data$created_date %<>% as.Date()
drug_shortages_data$anticipated_start_date %<>% as.Date()
drug_shortages_data$actual_start_date %<>% as.Date()
```

It's less elegant, but for the Drug Product Database (DPD), we have to download text files, but it's not so bad.

```{r download-from-dpd, echo=TRUE}
temp <- tempfile()

download.file(dpd_file_location, temp)

dpd_drug <- read_csv(
  file = unz(temp, "drug.txt"),
  col_names = colnames_drug,
  col_types = coltypes_drug)

dpd_status <- read_csv(
  file = unz(temp, "status.txt"),
  col_names = colnames_status,
  col_types = coltypes_status)

unlink(temp)
```

I'm also going to select a subset of the columns found in the data. The data contain `r drug_shortages_data %>% length()` columns.

I will select a small number of colunms ...

```{r select-columns}
drug_shortages_data %<>%
  select(din, created_date, anticipated_start_date, estimated_end_date,
         actual_start_date, actual_end_date, decision_reversal, updated_date,
         status, en_comments, en_drug_brand_name, en_drug_common_name,
         resolved, avoided, advance, hc_updated_date,
         en_supply_disruption_mitigation, en_discontinuation_comments,
         hc_en_comments, drug.current_status, drug.company.created_date,
         type.id, type.label)
```

... and subset the data to within the date range specified at the top of this document. I'm interested in the following date range: `r my_start_date` to `r my_end_date`.

```{r filter-by-dates}
drug_shortages_data_within_range <- drug_shortages_data %>%
  filter(ymd(created_date) %within% my_date_range)
```

<div class = "blue">

According to the code above, there were `r nrow(drug_shortages_data_within_range)` shortage reports between `r my_start_date` and `r my_end_date`.

</div>

<br>

## Filtering Data

Since I'm interested in the *percentage of actual drug shortage reports to Health Canada that did not have a prior report of an anticipated or actual shortage within a 12-month period for the same drug*, I'm going to filter through the data to find the subsets of interest. 

For instance I'd like to see the number of 

1. actual shortage reports
2. actual shortage reports that did not have a prior report of an anticipated or actual shortage within a 12-month period for the same drug

***

### Finding the Actual Shortages

I'm going to filter through the data I've pulled from the website and keep only the records that have a `status` of `active_confirmed`.

```{r filter-data-1}
actual_shortage <- drug_shortages_data_within_range %>%
  filter(status == "active_confirmed")

dins_for_actual_shortages <- actual_shortage$din %>% unique()
```

<div class = "blue">

According to the code above, there were `r nrow(actual_shortage)` actual shortages.

</div>

***

<br>

### Drug Shortages Without a Prior Report within a 12 Month Period

Next I'm going to keep only drug shortage reports to Health Canada that did not have a prior report of an anticipated or actual shortage within a 12-month period for the same drug. This means I'll take the data, keep only the drugs of interest from the previous step, keep only the records that had either an active or anticipated shortage and then drop anything that is more than 12 months old.

```{r filter-data-2}
more_than_12_months <- drug_shortages_data_within_range %>%
  filter(din %in% dins_for_actual_shortages) %>%
  filter(status %in% c("active_confirmed", "anticipated_shortage")) %>%
  filter(lubridate::today() - created_date > 365)
```

<div class = "blue">

According to the code above, there were `r nrow(more_than_12_months)` shortages more than 12 months old.

</div>

***

<br>

## Age on Market

```{r}

#' Mapping between DIN and DRUG_CODE:

din_drug_code_mapping <- dpd_drug %>% select(DRUG_CODE, DRUG_IDENTIFICATION_NUMBER)

#' Add DIN to table of interest:

my_drug_product_database_all_files_status <- merge(dpd_status,
                                                   din_drug_code_mapping,
                                                   by.y = "DRUG_CODE")


#' Ensure types are the same:
my_drug_product_database_all_files_status$DRUG_IDENTIFICATION_NUMBER %<>%
  as.character(my_drug_product_database_all_files_status$DRUG_IDENTIFICATION_NUMBER)

#' Drugs of interest:

drugs_of_interest <- more_than_12_months

drugs_of_interest$din %<>%
  as.character(drugs_of_interest$din) %>%
  str_pad(8, pad = "0")

my_drug_product_database_status <- merge(my_drug_product_database_all_files_status,
                                         drugs_of_interest,
                                         by.x = "DRUG_IDENTIFICATION_NUMBER",
                                         by.y = "din")


#' Find relevant data and calculate time on market:

my_drug_product_database_status_marketed <- my_drug_product_database_status %>%
  filter(STATUS == "MARKETED") %>%
  mutate(HISTORY_DATE = dmy(HISTORY_DATE)) %>%
  mutate(AGE_ON_MARKET = lubridate::today() - HISTORY_DATE)
```

<div class = "blue">

According to the code above, there were `r nrow(drug_shortages_data_within_range)` shortage reports between the dates of `r my_start_date` and `r my_end_date`. Of those reports, there were `r nrow(actual_shortage)` actual shortages.

Of the actual shortages, there were `r nrow(more_than_12_months)` shortages more than 12 months old.

</div>

<br>

### In Conclusion

<div class = "blue">

The proportion for the above is `r nrow(more_than_12_months) / nrow(actual_shortage)`.

<br>

The average age on the market for these drugs was `r mean(my_drug_product_database_status_marketed$AGE_ON_MARKET)` days.

</div>